---
title: "Lab2"
author: "Tristen Tooming"
date: "1/18/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### 1
The data set Hitters, which can be obtained from the ISLR library, contains data with 322 observations of major league baseball players on 20 variables that are assumed to influence the salary of the players.

The purpose of this assignment is to compare the results of

Best Subset Selection
Forward Stepwise Selection
Backward Stepwise Selection

Use the leaps library and set up models with all variables as predictors and Salary as response. Find and present the best models based on both BIC and Cp. Which method performs best based on the lowest BIC and Cp, and which variables are important for the salary for this model?

```{r}
library(ISLR)
library(leaps)

summary(Hitters)
```


##### Best Subset Selection
```{r}
# Fit
# Showing only best (nbest = 1)
regfit.models <- regsubsets(Salary~., data = Hitters, nbest = 1, nvmax = ncol(Hitters))

# Summary, Cp, BIC
res.sum <- summary(regfit.models)
as.data.frame(res.sum$outmat)

plot(regfit.models, scale='Cp')
plot(regfit.models, scale='bic')

cat(sprintf('Best Cp model: %s \n', which.min(res.sum$cp)))
cat('Coeffs for the model 10: \n\n')
coef(regfit.models, 10)

cat('\n\n')
cat(sprintf('Best BIC model: %s \n', which.min(res.sum$bic)))
cat('Coeffs for the model 6: \n\n')
coef(regfit.models, 6)
```

##### Forward Stepwise Selection

```{r}
regfit.fwd = regsubsets(Salary∼., data=Hitters, nvmax=ncol(Hitters), method ="forward ")
```


##### Backward Stepwise Selection
```{r}
regfit.bfd = regsubsets(Salary∼., data=Hitters, nvmax=ncol(Hitters), method ="forward ")
```

