---
title: "Lab1"
author: "Tristen Tooming"
date: "2/13/2021"
output: html_document
---

### 1
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#### A
```{r}
n <- 500 # Number of Observations (Rows)
p <- 20 # Number of "Variables"
ntr <- 5 # Number of datasets to be simulated
sig <- 0.2 # Scale parameter
X <- matrix(rnorm(n*p),ncol=p,nrow=n) # Generating values
set.seed(3) # Seed for the random generator
b_imp <- rnorm(ntr) # Random draw from normal distribution
b_zero <- rep(0,(p-ntr)) # zero vector
b_true <- c(b_imp,b_zero) #
y <- X%*%b_true + sig * rnorm(n) # Generating response values
```

#### B
```{r}
# Data frame
data = data.frame(y=y, X=X)
# Splitting data
train_portition = 0.7
sample_size = floor(n * train_portition)
set.seed(2707)
train_index = sample(seq_len(n), size = sample_size)
train = data[train_index, ]
test = data[-train_index, ]
```

#### C
```{r}
# Fitting the Model
linear_model = lm(y ~ ., data = train)
# Anova
anova(linear_model)
```

> Based on the F-test variables X.1, X.2, X.3, X.4 and X.5 are significant. Variable X.6 could be
useful for improving model performance but needs more validation than simple F-test.

```{r}
mse_train = c()
mse_test = c()
variables = names(data)[-1]
for (i in 1:p) {
  lm_formula = as.formula(paste('y ~ ', paste(variables[1:i], collapse = '+')))
  fit_train = lm(lm_formula, data=train)
  mse_train = c(mse_train, anova(fit_train)['Residuals', 'Mean Sq'])
  mse_test = c(mse_test, mean((test$y - predict.lm(fit_train, test)) ^ 2))
}
```
```{r}
plot(mse_train, type = 'l', col = 'red', ylab = 'MSE')
lines(mse_test, type = 'l', col = 'blue')
legend(15, 1.25, col = c('red', 'blue'), legend = c('Train', 'Test'), pch = 16)
```
```{r}
knitr::kable(data.frame(Iteration = 1:length(mse_test),MSE_Train = mse_train, MSE_Test = mse_test))
```

> Model performance increases significally when fourth term is included and reaches its maxium with fifth
predictor. After that there is no significant increase in performance when comparing models with MSE. We
choose to use model with formula y ~ X1 + X2 + X3 + X4 + X5. Minimum MSE test is at iteration #5 

### 2

```{r}
# Reading data
url = 'https://archive.ics.uci.edu/ml/machine-learning-databases/00236/seeds_dataset.txt'
wheat_seeds = read.table(url)
# Factoring and Naming variables
column_names = c('Area_A',
'Åerimeter_P',
'Compactness_C',
'Length_of_kernel',
'Width_of_kernel',
'Asymmetry_coefficient',
'Length_of_kernel_groove',
'Variety')
colnames(wheat_seeds) = column_names
variety_names = c()
wheat_seeds$Variety = as.factor(wheat_seeds$Variety)
set.seed(3)

wheat_seeds_permuted = sample(wheat_seeds)
# Splitting data
train_portition = 0.7
n = nrow(wheat_seeds)
sample_size = floor(n * train_portition)
set.seed(2707)
train_index = sample(seq_len(n), size = sample_size)
wheat_seeds_train = wheat_seeds_permuted[train_index, ]
wheat_seeds_test = wheat_seeds_permuted[-train_index, ]

calc_acc = function(actual, predicted) {
accuracy = sum( actual == predicted)/length(actual) * 100
return(round(accuracy, 2))
}
```


```{r}
# Packages
library(nnet)
# Model fit
model_multinom = multinom(Variety ~ ., data = wheat_seeds_train)
predicted_multinom_wheat_seeds = predict(model_multinom, wheat_seeds_test)
acc_multinom = calc_acc(wheat_seeds_test$Variety, predicted_multinom_wheat_seeds)
cat(sprintf("\n\nModel Acc: %s", acc_multinom))
```


```{r}
library(class)
accuracies = c() # Index = K-value
for (i in 1:25) {
fitted_knn = knn(train = wheat_seeds_train,
test = wheat_seeds_test,
k= i,
cl = wheat_seeds_train$Variety)
accuracies = c(accuracies, calc_acc(wheat_seeds_test$Variety, fitted_knn))
}
for (i in 1:length(accuracies)) {
cat(sprintf("K-value: %s, Accuracy: %s\n", i, accuracies[i]))
}


cat(sprintf('\n\nBest Model: %s', which(accuracies==max(accuracies))
))

acc_knn = max(accuracies)
```


#### C
```{r}
library(MASS)
model_lda = lda(Variety ~ ., data = wheat_seeds_train)
predicted_lda = predict(model_lda, wheat_seeds_test)$class
acc_lda = calc_acc(wheat_seeds_test$Variety, predicted_lda)
cat(sprintf('LDA accuracy %s', acc_lda))
```


```{r}
barplot(c(acc_multinom, acc_knn, acc_lda),
names.arg = c('Multinom', 'KNN', 'LDA'),
main = 'Prediction Accuracies',
col = c('#b2e8c2', '#33a892', '#483f72'))


pairs(wheat_seeds, col = wheat_seeds$Variety, main = "Matrix Scatterplot of the Data")
```


>Conclusion With 100% (K-value = 1) accuracy KNN performs best then comes LDA with 95% accuracy
and finally Multinomial with 89% accuracy. As seen in the scatter plot matrix varieties are forming distinct
6
clusters from each other and so on indicating good performance of the KNN. All in all I would choose KNN
for this particular dataset for predicting wheat varieties for its good accuracy, cheap computing costs and
specially for its simplicity
